---
title: "jake_summary_stats"
author: "Jake Birnbach"
date: "2024-04-26"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
original_par <- par()

#df <- read.csv("Data/data_weights.csv")
#dim(df)
#set.seed(1)
#df_sampled <- df %>% 
# sample_frac(0.001)

df_sampled <- read.csv("Data/data_sample_weights.csv")
df_sampled <- subset(df_sampled, select= -PERWT)
#write.csv(df_sampled, "Data/data_sample_weights.csv", row.names = FALSE)

n <- dim(df_sampled)[1]

df_sampled$SEX <- factor(df_sampled$SEX)
df_sampled$RACE <- factor(df_sampled$RACE)
df_sampled$DEGFIELD <- factor(df_sampled$DEGFIELD)
df_sampled$WorkRegion <- factor(df_sampled$WorkRegion)
```

A histogram of the outcome. You may use a kernel density plot instead, provided you
clearly explain which kernel you’re using. If your outcome is binary, you do not need
a histogram. Instead, just report the number and proportion of the observations that
had an outcome of 1. If you want to use a plot to do this, use a barplot
```{r}
hist(df_sampled$INCTOT, xlab="Total Income ($)", ylab="Frequency", main="Distribution of Total Personal Income")
```

Total personal income appears to be significantly right skewed which is expected.
Preforming a log transformation to make it normally distributed

```{r}
length(df_sampled$INCTOT[df_sampled$INCTOT <= 0])/n
df_sampled$INCTOT <- log(df_sampled$INCTOT)
df_sampled <- df_sampled[!is.na(df_sampled$INCTOT), ]

income = df_sampled$INCTOT
hist(df_sampled$INCTOT, xlab="Log Total Income ($)", ylab="Frequency", main="Distribution of Log Income", freq=FALSE, breaks=80)
curve(dnorm(x, mean(income), sd(income, na.rm = TRUE)), from=0, to=15, col='red', add=TRUE)
```

```{r}
summary(df_sampled)
```

Measures of central tendency (e.g., mean, median) and spread (e.g., standard deviation,
quartiles, range) of each variable. Categorical variables do not have central tendency
or spread. So, just report the number and proportion of observations in each category.
If you want to use a plot to do this, use a barplot
```{r}
par(original_par)

#SEX
summary(df_sampled$SEX)/n
plot(df_sampled$SEX, xlab="Sex", ylab="Count", main="Distribution of Men vs. Women")

#AGE
summary(df_sampled$AGE)
paste("Sdev Age: ",sd(df_sampled$AGE))
hist(df_sampled$AGE, xlab="Age (Years)", main="Distribution of Ages")


#RACE
par(mar=c(5, 12, 4, 2)+3)
race_table <- table(df_sampled$RACE)
summary(df_sampled$RACE)/n
barplot(sort(race_table), xlab="Count", main="Distribution of Races", horiz=TRUE, las=1)
par(original_par)


#UHSWORK
summary(df_sampled$UHRSWORK)
paste("Sdev UHRSWORK: ",sd(df_sampled$UHRSWORK))
hist(df_sampled$UHRSWORK, xlab="Hours Worked", main="Distribution of Hours Worked Per Week")

#WorkRegion
workregion_table <- table(df_sampled$WorkRegion)  # Create a frequency table

summary(df_sampled$WorkRegion)/n
barplot(sort(workregion_table), xlab="Count", main="Distribution of Working Regions", horiz=TRUE, las=1)

#DEGFIELD
par(mar=c(5, 12, 4, 2)+5)
summary(df_sampled$DEGFIELD)/n
plot(df_sampled$DEGFIELD, xlab="Degree Field", ylab="Frequency", main="Distribution of Degree Fields")
par(mar=c(5, 12, 4, 2)+3)


# Assuming df_sampled$DEGFIELD is a factor or can be treated as categorical data
degree_field_table <- table(df_sampled$DEGFIELD)  # Create a frequency table

# Adjust margins (bottom, left, top, right in number of lines of text)
par(mar=c(5, 12, 4, 2))

# Create a barplot
barplot(sort(degree_field_table), 
        main="Distribution of Degree Fields", 
        xlab="Count",
        las=1,
        horiz=TRUE)  # Setting las=2 makes the labels perpendicular to the axis

par(original_par)

```



A scatterplot matrix of your variables. The R function pairs will create this.
```{r}
pairs(df_sampled)
```

